\pkg{sparseHessianFD}: An R Package for Estimating Sparse Hessian Matrices

### 2016-12-22 nobuild/jss_revision/response.tex

\newcolumntype{d}[1]{D{.}{.}{#1}}
\newcommand{\pkg}[1]{\emph{#1}}
\newcommand{\proglang}[1]{\textsf{#1}}
\newcommand{\filename}[1]{\textit{#1}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\func}[1]{\code{#1}}
\newcommand{\class}[1]{\textsl{#1}}
\newcommand{\funcarg}[1]{\code{#1}}
\newcommand{\variable}[1]{\code{#1}}
\newcommand{\method}[1]{\func{#1}}

 \newcommand{\df}[3]{\mathsf{d}^{#1}f(#2;#3)}
 \newcommand{\parD}[3]{\mathsf{D}^{#1}_{#2}#3}
\newcommand{\hess}[2]{\mathsf{H}_{#1}#2}
\newcommand{\hessLT}[2]{\mathsf{L}_{#1}#2}
\newcommand{\Mat}[1]{\mathbf{#1}}
\newcommand{\Real}[1]{\mathbb{R}^{#1}}

\newenvironment{revQuote}{\itshape}{\vspace{\baselineskip}}
\newenvironment{response}{\normalfont}{\vspace{\baselineskip}}

\usepackage{geometry}
\geometry{margin=1.1in}

\setlength{\parskip}{1ex}
\setlength{\parindent}{0ex}

@@

\begin{document}

{\bfseries
\begin{center}
  Manuscript JSS2717\\
 ``sparseHessianFD: An R Package for Estimating Sparse
 Hessian Matrices''\\
 \vspace{\baselineskip}
 Response to reviewer
\end{center}
}
\vspace{\baselineskip}

I would like to express my sincere appreciation for the time and effort that
you invested in the review.  It is obvious to me that you went through
the paper and code in great detail.  In
particular, thank you for bringing the complex step
method to my attention.  I \emph{never} use complex numbers or
functions in my own work, so I typically ignore anything with the word
``complex'' in it.  But this review prompted me to learn something
new, and that is always fun.

I am happy to write that I was able to incorporate a large majority of
your suggestions into the the revised paper and package. For the
others, I hope that you, and the editor, will accept my
explanations and excuses.  Of course, if there are any strong
objections to how I responded to any of the comments, I am happy
to make further changes in advance of publication.

Below are my responses to each item, in the same order as in the
review.  The original comment is in italics, and is followed by my
response in normal font.  To save space, I removed the code and output
that were included in the review.

@@

{\normalfont (error messages related to algorithm.sty removed).}

I don't remember if there is an easy way to distribute ``special'' sty files,
or if you must force the user to install them. If the sty is not really
critical it might be better to omit it.

\end{revQuote}

\begin{response}
I suspect that the error is a result of the \pkg{algorithm} package
not being present in your LaTeX installation.  It is a standard
package in the MacTeX distribution, but I cannot speak with authority about other
platforms. It should be easily installed through CTAN.

I would prefer to continue to use the \pkg{algorithm} package, but I
could try to find another way to format the algorithms if there is a
strong objection.
\end{response}

@@

\end{response}

\item \begin{revQuote}
It is pointed out (p2) that requirements are burdensome and emphasized
that the package is not appropriate for all uses. Also, the conditions 1-5
are fairly stringent. The reader is left with the impression that there may
be few applications. I think the value of the package could be promoted
better. Perhaps a small list of different types or applications of
hierarchical problems could be given, or an indication of other classes of
problems where the conditions would apply.

  \end{revQuote}

  \begin{response}
I included this section primarily as a way to avoid overselling the
method, but the point that the reader might be scared away by this list of
restrictions is well-taken.  In the revision, I removed the list, and
wove the restrictions into the text.  The revision projects a more
positive tone that in the original submission.
\end{response}

@@

\item \begin{revQuote}
Condition 5 might be relaxed to a local condition in a neighborhood of
the evaluation point, rather than a global condition, but I do not know if
that would have any practical value.
  \end{revQuote}

\begin{response}
  The list of conditions is no longer in the paper, but to this point, the
  condition is global and not local.  I added a defintion of a ``structural zero'' in the third
  paragraph. The sparsity pattern represents the structure of the
  Hessian, not its value, so any element that \emph{could} be zero,
  but does not have to be zero, is not a structural zero.

   At the end of Section 2.2, I added a paragraph that explains that
   if there is uncertainty about whether an element is a structural
   zero or not, it is better to include it in the sparsity pattern.
\end{response}

\item \begin{revQuote}
Section 2.1 p7 I think could be made more easily readable by adding a few
more hints about dimension, and more care about use of the term 
"coefficients vector", for example:
\begin{itemize}
  \item 'continuous covariates $x_i$,' -> 'continuous covariates $x_i \in
\Real{k}$,'

  \item  'heterogeneous coefficient vector $\beta_i$' -> 'heterogeneous
coefficient vector $\beta_i \in \Real{k}$'

  \item 'The coefficients are distributed' -> 'The coefficient vectors
$\beta_i$ are distributed'
\end{itemize}
  \end{revQuote}

\begin{response}
  Thank you for the suggestions.  Changes have been made throughout the paper.
\end{response}

@@

\item \begin{revQuote}
p7 following "the cross-partial derivatives"
$\hess{\beta_i,\beta_j}{}=\parD{2}{\beta_i, \beta_j}{}=0$ for all $i\neq
j$.

  Is this a definition of $\hess{\beta_i,\beta_j}{}$ in terms of  $\parD{2}{}{}$
or a statement of equality?

  Also, I am confused by the single subscript $\beta_i$ to \\hess here, but a
double subscript just above in "Thus, $\hess{\beta_{ik},\mu_k}{}\neq 0$".
  \end{revQuote}

\begin{response}
  Based on this comment, I rewrote Section 2.2 to distinguish the
  cross-partial derivatives $\parD{2}{}{}$ from the corresponding
  elements of the Hessian $\hess{}{}$.  I did this by defining $x_i$ as
  a subset of the elements of $x$, and Ind$(x_i)$ as an 
  ``indexing'' function.  This change required me to redefine the
  covariate matrix in the example from $\Mat{X}$ to $\Mat{Z}$.

  The revised version is slightly more formal
  and precise, and I hope that this is the kind of change you had in
  mind.
\end{response}

\item \begin{revQuote}
 I think possibly (16) is ``banded'' and (17) ``block arrow'' rather than the
reverse which is indicated.
\end{revQuote}

\begin{response}
The descriptions of the two sparsity patterns are correct. The
first index in the subscript of $\beta$ refers to a household, and the
second index refers to a covariate. Thus,
$\parD{2}{\beta_{11},\beta_{12}}{}\neq 0$, but
$\parD{2}{\beta_{11},\beta_{21}}{}=0$.  When the second index changes
faster than the first, the pattern is ``block-arrow.''  When the first
index changes faster than the second, the pattern is ``banded.''

@@

\end{response}

\item \begin{revQuote}
Code in the paper prior to table 4 needs to be copied from the
vinettes/sparseHessianFD.Rnw file. For those trying to reproduce results it
would be nice if this where mentioned in the file replication.R.
  \end{revQuote}

\begin{response}
  Note to me:  I think what this means is that results for Table 4 and
  Figure 4 are "hard coded.''  For the jss article, the replication
  file is provided, but this file is not part of the package.  It
  should be, so non-jss vignette readers know how to replicate the results. 
\end{response}

@@

On the other hand, all.equal(hs, true.hess) is comparing a true analytic
calculation with a first order simple difference aproximation using the true
gradient function:

(Code removed)

which might also be mentioned in the text. (Really just for exposition
purposes, after all, it is almost the main purpose of the package.)
  \end{revQuote}

\begin{response}
  Good point.  I revised that section to explain that evaluations of
  the function and gradient, as called from the
  \class{sparseHessianFD} object, have to be identical to the true values.
  I replaced \func{all.equal} with \func{identical} to
  highlight that fact. I then explain that the
  \class{sparseHessianFD} calculation of the Hessian will not be
  identical to the true value, and report the mean relative
  difference.  I chose the mean relative difference over maximum
  absolute difference to be consistent with what \func{all.equal}
  uses to test if two matrices are equal within numeric tolerance.
\end{response}

\item \label{it:complex1}\begin{revQuote}
I think it would be instructive to add some of the following comparisons
with the above on p12.  The package numDeriv function hessian by default
does a second order Richardson approximation using the true function value
approximation. This involves a very large number of function evaluations in
an attempt to obtain some accuracy, but the accuracy is limited by being an
approximation of a second difference:

(Code removed)

@@

  \end{revQuote}

\begin{response}

Here I am combining my responses to items \ref{it:complex1} and
\ref{it:complex2} together.

  The complex step function is now implemented in the package, and is
  introduced in Section 2.5 of the paper.  You
  are correct that ``most R primitives work with complex numbers,''
  but there are many commonly used functions that do not, including 
  \func{log1p}, \func{gamma}, and the probability distribution
  functions. This limitation, combined with the requirement that the function be
  holomorphic, and my own inexperience with the complex step method, make
  me hesitant to promote it too heavily in the paper. But it is a nifty
  trick, and I am happy that it is now included in the package.

In terms of accuracy, the paper now includes comparisons between the true
Hessian and the two \class{sparseHessianFD} methods (finite differences
and complex step). I am not comparing accuracy with \pkg{numDeriv},
because that really is not the point of this paper.  However, for the
timing comparisons in Table 4, I did
change the baseline \pkg{numDeriv} method from \func{hessian} with the
\method{Richardson} method, to \func{jacobian} with the
\method{complex} method. Removing the second-order approximation saves
a lot of time, and lets me compute Table 4 without resorting to
parallel computation. That will simplify attempts at replication, and
removes another variable that might influence results.

I did collect timings using the \pkg{numDeriv} \method{simple} method,
but did not include those in Table 4 to save space.  There was not an
appreciable difference in computation time between the \method{simple}
and \method{complex} methods.

Footnote 1 was expanded to acknowledge the complex step method.

@@

  \end{revQuote}

\begin{response}
  I added a \func{set.seed} statement near the top of \filename{replication.R}.
\end{response}

\item \begin{revQuote}
Table 4 and 5. Some OS, processor, and memory details are helpful to put
timing results in context.
  \end{revQuote}

\begin{response}
  This information is now included in the first paragraph of Section 4.
\end{response}

@@

\item\begin{revQuote}
Table 4 is really not the proper comparison. I think a comparison with
       numDeriv::jacobian( binary.grad, P, method="simple", 
                   data=binary, priors=priors,
order.row=order.row)

really serves to highlight the improvement of the sparse calculation because
it is a valid comparison. Even though the results are not as exaggerated,
they are still important:

(My laptop is a  Intel(R) Core(TM) i5-3337U CPU @ 1.80GHz,  4GB RAM, SSD
swap, running Mint variant of Ubuntu 14.04.2 LTS.)

(Code and output removed)
  \end{revQuote}

  \begin{response}
    I agree.  See my response to Item \ref{it:complex2}.  I include the
    times from the \pkg{numDeriv}
    \method{complex} method instead of \method{simple} because of the
    improved accuracy.  I considered including results from both methods,
    but that would have cluttered the table, there was very
    little difference in the times between them, and the comparison
    between two \pkg{numDeriv} methods is not the point of this
    paper. But I completely agree that either \func{jacobian} method
    is a more valid comparison than using \func{hessian}.
  \end{response}

(On my laptop the next took about 30 hrs of which 24 was for the last,
N=2500, k=8 comparison.)

(Code and output removed)

@@

(Code and output removed)

  \end{revQuote}

  \begin{response}
Because the \func{numDeriv::jacobian} function runs so much faster
than  \func{numDeriv::hessian}, I was able to increase the size of the
examples in Table 4. Please forgive me for not going as high as
$N=2500$, $k=8$. One objective of Table 4 is to compare the relative
increases in computation time for different $N$ and $k$ between
\pkg{numDeriv} and \pkg{sparseHessianFD}, and I think that objective
is met with $N=(15, 50, 100, 500)$.  For the reader who wants to
replicate the results, a run time of more than a day is a lot to ask.
  \end{response}

@@

  \end{revQuote}

  \begin{response}
    Fixed.  Thanks.
  \end{response}

\item\begin{revQuote}

While that package is useful, and reasonably demonstrated in the paper,
I think it would be nice to expand the paper in some ways that might be
deemed more "original research". Some possibilities are:
\begin{itemize}
 \item Explain and try to assess how much of the speedup is due to simple
sparseness and how much is due to the "sparse patern" (p2) allowing for
perturbing multiple elements together. (I think these are related but
slightly different?)

 \item Try to assess how much of the speedup is due to reduce computing demand
and how much is due to different memory demand. (I had the impression in the
larger problems with numDeriv that my computer started to use swap space,
which resulted in a big slow down.)

 \item Consider implementing a complex step method, and do a comparison.

 \item  Assess the difference when multiple CPUs are used. (run.par == FALSE vs
run.par == TRUE)
\end{itemize}
  \end{revQuote}

  \begin{response}
These are all reasonable suggestions.  Let me address them point by point.
    \begin{itemize}
    \item It really is all about the sparsity pattern, and
      specifically, the how few partitions are needed to group the
      variables.  Of course, it will be hard to keep the number of
      partitions small if there are too many non-zero elements, so the
      sparsity itself has some effect.  The arrangement of the non-zeros is
      related to the ordering of the variables, but the permutation
      handles that during the initialization of the
      \class{sparseHessianFD} argument.  For example, variables
      arranged in a way that the Hessian has a block-arrow pattern are
      permuted to get a Hessian with a banded pattern instead.  This
      is all discussed in Section 3.1.

\item I have no idea how to do this in a generalizable way.  I think
  it has a lot to do with the architecture of the computer
  itself.  For example, my computer has 64 GB of RAM, so memory usage
  is not a binding constraint, and swap space never comes into play.

  Figure 4 does break down the components of the computation
  into the time it takes to reserve memory (initialization), partition
  the variables, and compute gradients themselves.  Perhaps that is
  the kind of comparison you had in mind.

@@

  I do not want to ignore this suggestion, but I am simply not clear on
  what is being asked.  I am open to more specific suggestions.

\item The package now implements the complex step method, and comparisons of
  accuracy and computation time are included in the paper.

@@

\item Parallel computing on a shared memory architecture really
  muddles up timing comparisons.  The \emph{only} reason I generated
  Table 4 in parallel was because of the \pkg{numDeriv} \func{hessian}
  function. Now that \func{hessian} was replaced with \func{jacobian},
  the baseline measures from \pkg{numDeriv} run fast enough that I can
  generate Table 4 in serial.

Your point that timing comparisons can be affected by
whether computation is being run on multiple processors is well taken, but the
``serialization'' of Table 4 makes the issue moot.

  Data for Figure 4 were generated in parallel because of the large
  number of required simulations.

### 2016-12-22 nobuild/jss_revision/sparseHessianFD.Rnw
more on revision

\date{December 20, 2016}

@@

The \pkg{sparseHessianFD} package is a tool for estimating
sparse Hessians using finite differences of gradients.  Section \ref{sec:numdiff} will
cover the specifics, but the basic idea is as follows.  Consider a
function $f(x)$, its gradient $\nabla{f(x)}$, and its Hessian
$\hess{}{f(x)}$, for $x\in\mathbb{R}^M$.  Define the derivative vector
as the transpose of the gradient, and a vector of partial derivatives, so
${\parD{}{}{f(x)}=\nabla f(x)^\top=\left(\parD{}{1}{},\mathellipsis,\parD{}{M}{}\right)}$.   (Throughout the paper, we will try
to reduce notational clutter by referring to the derivative and
Hessian as $\parD{}{}{}$ and $\hess{}{}$, respectively, without the
$f(x)$ symbol).  Let $e_m$ be a vector of zeros, except with a 1 in the $m$th
element, and let $\delta$ be a sufficiently small scalar
constant. A linear approximation to the $m$th column of the Hessian is ${\hess{m}{}\approx\left(\nabla f(x+\delta
  e_m)- \nabla f(x)\right)/\delta}$.  Estimating a dense Hessian in this way involves at least $M+1$ calculations of the
gradient: one for the gradient at $x$, and one after perturbing each
of the $M$ elements of $x$, one at a time.\footnote{More accurate
  approximations require more gradient evaluations.  We will consider
  only forward finite differences.}  In a sparse matrix, most of the
elements are constrained to zero.  Depending
on the sparsity pattern of the Hessian, those constraints may let us recover
the Hessian with fewer gradient evaluations by perturbing multiple
elements of $x$ together.  For some sparsity patterns, estimating a Hessian in
this way can be profoundly efficient.  In fact, for the hierarchical models that we consider in
this paper, the number of gradient evaluations does not increase with
the number of additional heterogeneous units.

@@

From the log posterior density in Equation \ref{eq:LPD}, we can see
that $\parD{2}{\beta_i, \beta_i}\neq 0$ (one element of $\beta_i$ could be
correlated with another element of $\beta_i$), and that, for all $i$,
$\parD{2}{\beta_i, \mu}=0$ (because $\mu$ is the prior mean of each
$\beta_i$).  However, since the $\beta_i$ and $\beta_j$ are independently distributed, and the $y_i$ are
conditionally independent, the cross-partial derivatives
$\parD{2}{\beta_i, \beta_j}=0$ for all $i\neq j$. When $N$ is much
greater than $k$, there will be many more zero cross-partial
derivatives than non-zero.  Regardless of how each $\parD{2}{}$ is
mapped to $\hess{}{}$ through the indexing function, the Hessian of the log
posterior density will be sparse.

@@

The sparsity pattern required by \pkg{sparseHessianFD} consists of the row and column
indices of the non-zero elements in the \emph{lower triangle} the
Hessian, and it is the responsibility of the user to ensure that
the pattern is correct.  In practice, rather than trying to keep track of the row and column indices
directly, it might be easier to construct a pattern matrix first,
check visually that the matrix has the right pattern, and then extract
the indices.

The package defines utility functions (Table~\ref{tab:helpers}) to convert
between sparse matrices, and the vectors of row and column indices
required by the \class{sparseHessianFD} initializer.

@@

@@

Finally, we create an instance of a \class{sparseHessianFD} object.
Evaluations of the function and gradient will generate the same
results as the true values because they are computed using the same function.

@@

Note that variables \variable{f} and \variable{true.f} were computed
using the same exact calculation, as were variables
\variable{gr} and \variable{true.grad}.In contrast,
\variable{true.hess} was computed using an exact calculation, while
\variable{hs} is the sparse finite difference approximation.

\subsection[Complex step method]{Complex step method}

@@

As far as we know, \pkg{numDeriv} \citep{R_numDeriv} is the only other \proglang{R} package
that computes numerical approximations to derivatives.  It differs from
\pkg{sparseHessianFD} in some important ways.

\begin{enumerate}
\item It treats all Hessians as dense;
  \item It computes each element of the Hessian using a second-order
    finite differencing approximation that does not require the user
    to supply the gradient; and
    \item It implements iterative algorithms to improve accuracy, at
      the expense of speed.
\end{enumerate}

Nevertheless, \pkg{numDeriv} is an easy-to-use tool for numerical differentiation,
so it is worthwhile to compare its performance to that of
\pkg{sparseHessianFD}.  To prepare Table~\ref{tab:numDeriv}, we
estimated Hessians of the log posterior density in Equation~\ref{eq:LPD} with different numbers of heterogeneous
units ($N$) and within-unit parameters ($k$).  As in the previous
section, the total number of
variables is $M=(N+1)k$.  Table~\ref{tab:numDeriv} shows the mean
and standard deviations (across 20 replications) for the time (in
milliseconds) to compute a Hessian using each package.  The difference in run times is dramatic,
especially because the computation time for \pkg{numDeriv} grows
quadratically in the number of variables.  The setup time for
\pkg{sparseHessianFD} was about 7 milliseconds for all cases.  Times
were collected using the \pkg{microbenchmark} package \citep{R_microbenchmark}.

@@

Because \pkg{numDeriv} does not scale, we cannot use it to benchmark \pkg{sparseHessianFD} for
larger sparse Hessians.  To help us understand just how scalable
\pkg{sparseHessianFD} is, we ran another set of simulations, for the
same hierarchical model, for different values of $N$ and $k$.  We then
recorded the run times for different steps in the sparse Hessian
estimation, across 200 replications.  The steps are summarized in Table~\ref{tab:timeMeasures}.

>>>>

\date{December 22, 2016}

@@

The \pkg{sparseHessianFD} package is a tool for estimating
sparse Hessians using finite differences of gradients.  Section \ref{sec:numdiff} will
cover the specifics, but the basic idea is as follows.  Consider a
function $f(x)$, its gradient $\nabla{f(x)}$, and its Hessian
$\hess{}{f(x)}$, for $x\in\mathbb{R}^M$.  Define the derivative vector
as the transpose of the gradient, and a vector of partial derivatives, so
${\parD{}{}{f(x)}=\nabla f(x)^\top=\left(\parD{}{1}{},\mathellipsis,\parD{}{M}{}\right)}$.   (Throughout the paper, we will try
to reduce notational clutter by referring to the derivative and
Hessian as $\parD{}{}{}$ and $\hess{}{}$, respectively, without the
$f(x)$ symbol).  Let $e_m$ be a vector of zeros, except with a 1 in the $m$th
element, and let $\delta$ be a sufficiently small scalar
constant. A linear approximation to the $m$th column of the Hessian is ${\hess{m}{}\approx\left(\nabla f(x+\delta
  e_m)- \nabla f(x)\right)/\delta}$.  Estimating a dense Hessian in this way involves at least $M+1$ calculations of the
gradient: one for the gradient at $x$, and one after perturbing each
of the $M$ elements of $x$, one at a time.\footnote{More accurate
  approximations using finite differences require more gradient evaluations.  We will consider
  only forward finite differences. In Section \ref{sec:complex} we
  introduce the complex step method that, under certain conditions,
  can generate more accurate estimates with the same number of
  gradient evaluations using forward finite differences.} In a sparse matrix, most of the
elements are constrained to zero.  Depending
on the sparsity pattern of the Hessian, those constraints may let us recover
the Hessian with fewer gradient evaluations by perturbing multiple
elements of $x$ together.  For some sparsity patterns, estimating a Hessian in
this way can be profoundly efficient.  In fact, for the hierarchical models that we consider in
this paper, the number of gradient evaluations does not increase with
the number of additional heterogeneous units.

@@

From the log posterior density in Equation \ref{eq:LPD}, we can see
that $\parD{2}{\beta_i, \beta_i}\neq 0$ (one element of $\beta_i$ could be
correlated with another element of $\beta_i$), and that, for all $i$,
$\parD{2}{\beta_i, \mu}\neq 0$ (because $\mu$ is the prior mean of each
$\beta_i$).  However, since the $\beta_i$ and $\beta_j$ are independently distributed, and the $y_i$ are
conditionally independent, the cross-partial derivatives
$\parD{2}{\beta_i, \beta_j}=0$ for all $i\neq j$. When $N$ is much
greater than $k$, there will be many more zero cross-partial
derivatives than non-zero.  Regardless of how each $\parD{2}{}$ is
mapped to $\hess{}{}$ through the indexing function, the Hessian of the log
posterior density will be sparse.

@@

The sparsity pattern required by \pkg{sparseHessianFD} consists of the row and column
indices of the non-zero elements in the \emph{lower triangle} the
Hessian, and it is the responsibility of the user to ensure that
the pattern is correct.  In practice, rather than trying to keep track of the row and column indices
directly, it might be easier to construct a pattern matrix first,
check visually that the matrix has the right pattern, and then extract
the indices.  The package defines utility functions (Table~\ref{tab:helpers}) to convert
between sparse matrices, and the vectors of row and column indices
required by the \class{sparseHessianFD} initializer.

@@

If there is uncertainty about whether an element is a structural
   zero or not, one should err on the side of it being non-zero, and
   include that element in the sparsity pattern.
   There might be a loss of efficiency if the element really is a
   structural zero, but the result will still be correct.  All that
   would happen is that the numerical estimate for that element would be zero
   (within machine precision). On the other hand, excluding a non-zero
   element from the sparsity pattern will likely lead to an incorrect
   estimate of the Hessian.

@@

Finally, we create an instance of a \class{sparseHessianFD} object.
Evaluations of the function and gradient using the \method{fn} and
\method{gr} methods will always give the same
results as the true values because they are computed using the same functions.

@@

\subsection[Complex step method]{Complex step method}\label{sec:complex}

@@

As far as we know, \pkg{numDeriv} \citep{R_numDeriv} is the only other \proglang{R} package
that computes numerical approximations to derivatives.  Like
\pkg{sparseHessianFD} it includes functions to compute Hessians from
user-supplied gradients (through the \func{jacobian} function), and
implements both the finite differencing and complex step methods.  Its
most important distinction from \pkg{sparseHessianFD} is that it treats all Hessians as dense.  Thus, we
will use  \pkg{numDeriv} as the baseline against which we can compare the performance of
\pkg{sparseHessianFD}.  All of the subsequent speed and scalability
results were generated on a 2013-vintage Apple Mac Pro with a 2.7 GHz
12-core Intel Xeon E5 processor and 64 GB of RAM.

To prepare Table~\ref{tab:numDeriv}, we
estimated Hessians of the log posterior density in Equation~\ref{eq:LPD} with different numbers of heterogeneous
units ($N$) and within-unit parameters ($k$).  As in the previous
section, the total number of
variables is $M=(N+1)k$.  Table~\ref{tab:numDeriv} shows the mean
and standard deviations (across 20 replications) for the time (in
milliseconds) to compute a Hessian using each package.  The difference in run times is dramatic,
especially because the computation time for \pkg{numDeriv} grows
quadratically in the number of variables.  The setup time for
\pkg{sparseHessianFD} was about 7 milliseconds for all cases.  Times
were collected using the \pkg{microbenchmark} package \citep{R_microbenchmark}.

@@

To help us understand just how scalable
\pkg{sparseHessianFD} is, we ran another set of simulations, for the
same hierarchical model, for different values of $N$ and $k$.  We then
recorded the run times for different steps in the sparse Hessian
estimation, across 200 replications.  The steps are summarized in Table~\ref{tab:timeMeasures}.

### 2016-12-22 nobuild/jss_revision/sparseHessianFD.tex
more on revision

\begin{Schunk}
\begin{Sinput}
R> hs <- obj$hessian(P)
R> mean(abs(hs-true.hess))/mean(abs(hs))
\end{Sinput}
\begin{Soutput}
[1] 1.12e-08
\end{Soutput}
\end{Schunk}

Note that variables \variable{f} and \variable{true.f} were computed
using the same exact calculation, as were variables
\variable{gr} and \variable{true.grad}.In contrast,
\variable{true.hess} was computed using an exact calculation, while
\variable{hs} is the sparse finite difference approximation.

\subsection[Complex step method]{Complex step method}

@@

\begin{Schunk}
\begin{Sinput}
R> obj2 <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
+   rows=pattern[["rows"]], cols=pattern[["cols"]],
+   complex=TRUE,
+   data=binary, priors=priors, order.row=order.row)
R> hs2 <- obj2$hessian(P)
R> mean(abs(hs2-true.hess))/mean(abs(hs2))
\end{Sinput}
\begin{Soutput}
[1] 6.75e-18
\end{Soutput}
\end{Schunk}

>>>>

\begin{Schunk}
\begin{Sinput}
R> hs <- obj$hessian(P)
R> mean(abs(hs-true.hess))/mean(abs(hs))
\end{Sinput}
\begin{Soutput}
[1] 1.1217e-08
\end{Soutput}
\end{Schunk}

\subsection[Complex step method]{Complex step method}\label{sec:complex}

@@

\begin{Schunk}
\begin{Sinput}
R> obj2 <- sparseHessianFD(P, fn=binary.f, gr=binary.grad,
+   rows=pattern[["rows"]], cols=pattern[["cols"]],
+   complex=TRUE,
+   data=binary, priors=priors, order.row=order.row)
R> hs2 <- obj2$hessian(P)
R> mean(abs(hs2-true.hess))/mean(abs(hs2))
\end{Sinput}
\begin{Soutput}
[1] 6.7543e-18
\end{Soutput}
\end{Schunk}
